---
title: "Data inputs for the SEFRA seabird model"
author: "Charles Edwards, Tom Peatman, Dave Goad, Darcy Webber"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    keep_tex: true
    number_sections: true
    citation_package: natbib
vignette: >
  %\VignetteIndexEntry{model_inputs}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
header-includes: 
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
  \usepackage{caption}
  \usepackage{lmodern}
  \renewcommand{\familydefault}{\sfdefault}
  \usepackage{sansmath}
  \sansmath
  \usepackage{float}
  \floatplacement{figure}{H}
bibliography: lib.bib
biblio-style: apa
---

```{r, include = FALSE}
library(knitr)
library(pander)
library(dplyr)
library(ggplot2)
```

```{r, include = FALSE}
options(encoding = "UTF-8")
opts_chunk$set(fig.width = 10, fig.height = 5.9, echo = FALSE, warning = FALSE, message = FALSE, tidy = 'styler', tidy.opts=list(strict = TRUE))

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

knitr::opts_knit$set(output.dir = ".")
```

```{r}
fig_nums <- captioner::captioner(prefix = "Figure")
tab_nums <- captioner::captioner(prefix = "Table")
```

\vfill
```{r, echo = TRUE, message = FALSE}
library(sraInputs)

data("species")
data("species_groups")
data("capture_effort")
```

### notes

plot bubble plot of observed and total overlap per species group and fishery group (with different color per method). make observed overlap bubble within total overlap to indicate coverage. 

figure 3: legend should say observed captures...

overlap tables: remove total overlap because it is not helpful

new table showing new and old overlap calculation per species only for species that have had new maps

overlap figure maps: can we have a line/violin plot per method showing distribution of overlap across months?

capture tables: need to distinguish between net and warp captures. can we include net/warp captures in map with different colors?

years for fishery groups: data are lost as we go back in time.

estimate catchability across all years; but captures for last 3 years only

recode fishery group definitions to include year time frame

flag instances in which overlap is zero but there are captures







```{r, echo = TRUE, message = FALSE}
library(sraInputsBio)

data("inputsBio_v1.0.0")
```

```{r}
# check that package data and version match
stopifnot(isTRUE(all.equal(attributes(inputsBio_v1.0.0)$version, packageVersion("sraInputsBio"))))
```

```{r}
# get data
inputsBio <- get("inputsBio_v1.0.0")
rm(inputsBio_v1.0.0)
```

```{r}
# fishing_years
fishing_years   <- paste0(2006:2018, "/", 2007:2019)
n_fishing_years <- length(fishing_years)
```

```{r}
# construct methods data frame
methods <- data.frame(id_method = 1:6, 
                      method = c("BLL", "SLL", "SN", "Trawl", "PS", as.character(NA)), 
                      include = c(rep(TRUE, 4), rep(FALSE, 2)))
```

```{r}
# fishing groups
source("fishing_group_functions.R")
fishing_groups <- fishing_groups_option_II
```

\newpage
# Inputs summary
The model includes `r n_fishing_years` years of data between `r min(fishing_years)` and `r max(fishing_years)`. A full list of species and their associated species groups is given in `r tab_nums("tab:species", display = "cite")`. Biological and fisheries data are sourced from the R-packages \href{https://www.github.com/cttedwards/sraInputsBio}{sraInputsBio} (version `r packageVersion("sraInputsBio")`) and \href{https://www.github.com/cttedwards/sraInputs}{sraInputs} (version `r packageVersion("sraInputs")`), respectively. 

Fishing methods included in the model were: `r unique(methods$method[methods$include])`, with associated fishing groups listed in `r tab_nums("tab:fishing_groups", display = "cite")`. Fishing groups have been updated from previous risk assessments [@Richard2017sra,@Richard2020sra] and were redefined based knowledge of fishing practices and chosen to reflect differences in the catchability of seabirds associated with fishing. 

Trawl groups were simplified and defined using a combination of length, target species, and presence of a freezer (Figure \ref{fig:trawlv1}). Bottom longline groups were updated to separate out small autoline vessels, and these were further split by target. Manual longline vessels were split by target species (Figure \ref{fig:bllv1}). Surface longline effort was split into the large vessel joint venture fleet and the smaller vessel domestic fleet (targetting tuna and swordfish). Further refinement of the fishery groups will be considered. For example, trawl fleets could be split by discard type, and some proposed bottom longline groups could be joined.

```{=tex}
\begin{figure}
\centering
\includegraphics[width = 0.7\textwidth]{trawlv1.eps}
\caption{Decision tree for partition of trawl groups.}
\label{fig:trawlv1}
\end{figure}
```

```{=tex}
\begin{figure}
\centering
\includegraphics[width = 0.7\textwidth]{bllv1.eps}
\caption{Decision tree for partition of BLL groups.}
\label{fig:bllv1}
\end{figure}
```

```{=tex}
\captionsetup[table]{labelformat=empty}
\captionsetup[figure]{labelformat=empty}
```

Species groups are currently fixed according to previous seabird risk assessments, and will likely be updated. Cryptic mortality multipliers are not presented here, as they are still being defined.

Observed captures by species, species group, fishing group and method are shown in `r tab_nums("tab:captures1", display = "cite")`, `r tab_nums("tab:captures2", display = "cite")`, `r fig_nums("fig:captures1", display = "cite")`, `r fig_nums("fig:captures2", display = "cite")` and `r fig_nums("fig:captures3", display = "cite")`. The model assumes that captures are proportional to the density overlap (`r fig_nums("fig:overlap_o", display = "cite")`). The overlap provides a model input for fitting and prediction of captures, and is listed in `r tab_nums("tab:overlap_o", display = "cite")` and `r tab_nums("tab:overlap_t", display = "cite")`. When calculating the overlap, effort units are of a thousand hooks for bottom longline (BLL) and surface longline (SLL), km of net for set net (SN), and number of tows for trawl. Species specific maps of the summed density overlap and captures are also provided.

\clearpage
```{r, results = "asis", size = "scriptsize"}
tab_cap <- tab_nums(name = "tab:species", caption = paste0("Species (", length(species),") included in the model and their species (", length(unique(species_groups$id_species_group)),") groups."))

species_groups <- species_groups %>%
  mutate(common_name = gsub("'", "’", common_name))

tab_sg <- species_groups %>%
  select(id_species, code_sra, common_name, species_group)

species_labels <- stringr::str_replace_all(tab_sg$common_name, "[[:punct:]]", "")

pandoc.table(tab_sg,
             caption = tab_cap,
             row.names = FALSE, 
             split.table = Inf, 
             justify = 'left',
             col.names = c("ID", "Code", "Common name", "Species group"))
```

\clearpage
```{r}
# read file to obtain proportions in zone per species and month
sbp <- read.csv("species_breeding_months.csv")
sbp <- sbp %>% 
     dplyr::select(id_species, code_sra, breeding, nonbreeding) %>% 
     tidyr::pivot_longer(cols = c("breeding", "nonbreeding"), values_to = "month", names_to = "period") %>% tidyr::separate_rows(month) %>% 
     mutate(month = as.integer(month))
```

```{r}
# get matrix of breeding / non-breeding per species per month.
breeding <- sbp %>% filter(!is.na(month)) %>%
  mutate(breeding = case_when(period == "breeding" ~ 1L, period == "nonbreeding" ~ 0L, TRUE ~ as.integer(NA))) %>% 
  tidyr::pivot_wider(id_cols = id_species, names_from = month, values_from = breeding) 

breeding <- breeding[order(breeding$id_species),] %>% select(!id_species) %>% as.matrix()
```

\newpage\clearpage

```{r}
# subset data
dat <- capture_effort %>% filter(fishing_year %in% fishing_years) %>% filter(!is.na(month)) %>% filter(!(is.na(start_lat) | is.na(start_lon)))
```


```{r}
# identify fishery groups
dat <- dat %>% mutate(id_fishing_group_option_II = case_when(method == "Trawl" ~ trawl_groups_option_IIA(method, fishery, target, vessel_key, vessel_class, start_date, vessel_size, vessel_freezer),
                                                   method == "BLL" ~ bll_groups_option_II(method, fishery, target, vessel_key, vessel_class, start_date, vessel_length),
                                                   method == "SLL" ~ sll_groups_option_II(vessel_class, method),
                                                   method == "SN" ~ setnet_groups_option_II(method),
                                                   method == "PS" ~ purseseine_groups_option_II(method),
                                                   TRUE ~ as.numeric(NA))) %>% filter(!is.na(id_fishing_group_option_II))

# organise group names
fishing_groups <- fishing_groups_option_II %>% 
  left_join(methods, by = "method") %>% 
  filter(include, id_fishing_group_option_II %in% unique(dat$id_fishing_group_option_II)) %>% 
  select(-include) %>%
  mutate(id_fishing_group = 1:nrow(.))

# join
dat <- dat %>% left_join(fishing_groups, by = c("id_fishing_group_option_II", "method")) %>% select(-id_fishing_group_option_II) %>% rename(fishing_group = fishing_group_name)

# remove unwanted methods
dat <- dat %>% filter(!is.na(id_method))

methods <- methods %>% filter(include)
```

```{r, results = "asis", size = "scriptsize"}
tab_cap <- tab_nums(name = "tab:fishing_groups", caption = paste0("Fishing methods (", nrow(methods), ") and fishing groups (", nrow(fishing_groups), "). Observed and total effort, and percentage coverage, are reported using the number of fishing events."))

tab <- dat %>% group_by(id_fishing_group) %>% 
  summarise(method = unique(method), 
            id_method = unique(id_method),
            fishing_group = unique(fishing_group_label), 
            year_start = "2006/07",
            year_end = "2018/19",
            observer_effort = sum(observer_effort_num), 
            total_effort    = sum(effort_num),
            coverage_percent = 100 * round(observer_effort / total_effort, 3), 
            n_captures = sum(n_captures, na.rm = TRUE)) 

tab <- tab %>% select(method, fishing_group, id_fishing_group, year_start, year_end, observer_effort, total_effort, coverage_percent, n_captures)

tot <- c(method = "Total", fishing_group = NA, id_fishing_group = NA, year_start = NA, year_end = NA, colSums(tab[,6:9], na.rm = TRUE))
tot[8] <- NA 
tab <- rbind(tab, t(tot)) 

pandoc.table(tab,
             caption = tab_cap,
             justify = c("left", "left", "left", "left", "left", "right", "right", "right", "right"), 
             emphasize.strong.rows = nrow(tab),
             missing = "-",
             row.names = FALSE, 
             split.table = Inf,
             col.names = c("Method", "Fishing group", "ID", "Start", "End", "Observed effort", "Total effort", "Coverage (%)", "n captures"))
```

\clearpage
```{r, echo = FALSE, results = "asis", size = "scriptsize"}
tab_cap <- tab_nums(name = "tab:captures1", caption = "Number of observed seabird captures by species and method.")

tab <- dat %>% as.data.frame() %>%
  aggregate_captures("method", with(species_groups, split(code_sra, common_name))) %>% 
  tidyr::pivot_wider(names_from = method, values_from = n_captures, values_fill = list(n_captures = 0)) %>%
  mutate(Total = BLL + SLL + SN + Trawl)

#tab <- tab %>% left_join(species_groups, by = c("group" = "code_sra")) %>% select(common_name, BLL, SLL, SN, Trawl, Total)

tot <- data.frame(group = "Total", t(colSums(tab[,2:6], na.rm = TRUE)))
tot <- rbind(tot, data.frame(group = "Total per year", t(colSums(tab[,2:6], na.rm = TRUE) / n_fishing_years)))
tab <- rbind(tab, tot)

tab[tab == 0] <- NA

pandoc.table(tab,
             caption = tab_cap,
             row.names = FALSE, 
             emphasize.strong.rows = c(nrow(tab) - 1, nrow(tab)),
             emphasize.strong.cols = ncol(tab),
             missing = "-",
             split.table = Inf, 
             justify = c('left', rep('right', 5)), 
             col.names = c("Common name", names(tab)[2:ncol(tab)]))
```

\clearpage
```{r, echo = FALSE, results = "asis", size = "scriptsize"}
tab_cap <- tab_nums(name = "tab:captures2", caption = "Number of observed seabird captures by species group and method.")

tab <- dat %>% as.data.frame() %>%
  aggregate_captures("method", with(species_groups, split(code_sra, species_group))) %>% 
  tidyr::pivot_wider(names_from = method, values_from = n_captures, values_fill = list(n_captures = 0)) %>%
  mutate(Total = BLL + SLL + SN + Trawl, Rate = Total / n_fishing_years)
tab <- tab[rev(order(tab$Total)),]

tot <- data.frame(group = "Total", t(colSums(tab[,2:6], na.rm = TRUE)), Rate = NA)
tot <- rbind(tot, data.frame(group = "Rate per year", t(colSums(tab[,2:6], na.rm = TRUE) / n_fishing_years), Rate = NA))
tab <- rbind(tab, tot)

tab[tab == 0] <- NA

pandoc.table(tab,
             caption = tab_cap,
             row.names = FALSE, 
             emphasize.strong.rows = c(nrow(tab) - 1, nrow(tab)),
             emphasize.strong.cols = c(ncol(tab) - 1, ncol(tab)),
             missing = "-",
             split.table = Inf, 
             justify = c('left', rep('right', 6)), 
             col.names = c("Group name", names(tab)[2:6], "Rate per year"))
```

`r fig_cap <- fig_nums(name = "fig:captures1", caption = "Observed captures per fishery group.")`

```{r, fig.cap = fig_cap}
dfr <- dat %>% as.data.frame() %>%
  aggregate_captures(c("method", "fishing_group_label", "fishing_year"), species) %>% summarise(n_captures = sum(n_captures))

dfr$n_captures[dfr$n_captures == 0] <- NA
dfr$fishing_year  <- factor(dfr$fishing_year,        levels = fishing_years)
dfr$fishing_group <- factor(dfr$fishing_group_label, levels = unique(dfr$fishing_group_label))

ggplot(dfr) + 
  geom_point(aes(x = fishing_year, y = fishing_group, size = n_captures, col = method), alpha = 0.4) + 
  scale_size(range = c(1, 20), name = "Captures (N)") + 
  scale_x_discrete(breaks = levels(dfr$fishing_year)[c(2, 4, 6, 8, 10, 12)]) + 
  labs(x = "", y = "", col = "")
```

`r fig_cap <- fig_nums(name = "fig:captures2", caption = "Observed captures per species group.")`

```{r, fig.cap = fig_cap}
species_group_lookup <- with(species_groups, split(code_sra, species_group))

dfr <- dat %>% as.data.frame() %>%
  aggregate_captures("fishing_year", species_group_lookup)

dfr$n_captures[dfr$n_captures == 0] <- NA
dfr$fishing_year <- factor(dfr$fishing_year, levels = fishing_years)
dfr$group        <- factor(dfr$group, levels = rev(unique(species_groups$species_group)))

ggplot(dfr) + 
  geom_point(aes(x = fishing_year, y = group, size = n_captures), alpha = 0.4) + 
  scale_size(range = c(1, 20), name = "Captures (N)") + 
  scale_x_discrete(breaks = levels(dfr$fishing_year)[c(2, 4, 6, 8, 10, 12)]) + 
  labs(x = "", y = "", col = "")
```

`r fig_cap <- fig_nums(name = "fig:captures3", caption = "Observed captures per species and fishery group combination.")`

```{r, fig.cap = fig_cap, fig.height = 12}
dfr <- dat %>% as.data.frame() %>%
  aggregate_captures("fishing_group_label", species_group_lookup)

dfr$n_captures[dfr$n_captures == 0] <- NA
dfr$fishing_group <- factor(dfr$fishing_group_label, levels = unique(dfr$fishing_group_label))
dfr$group         <- factor(dfr$group, levels = rev(unique(species_groups$species_group)))

ggplot(dfr) + 
  geom_point(aes(x = fishing_group, y = group, size = n_captures), alpha = 0.4) + 
  scale_size(range = c(1, 20), name = "Captures (N)") + 
  theme(axis.text.x = element_text(angle = 90), axis.text = element_text(size = 16)) +
  labs(x = "", y = "", col = "")
```

\clearpage
```{r, results = "asis", size = "scriptsize"}
tab_cap <- tab_nums(name = "tab:effort_o", caption = paste("Observed fishing effort by method and fishing year between ", min(fishing_years)," and ", max(fishing_years), ". Units are of a thousand hooks for bottom longline (BLL) and surface longline (SLL), km of net for set net (SN), and number of tows for trawl."))

dfr <- dat %>% as.data.frame() %>%
  group_by_at(c("fishing_year", "method")) %>% summarise(observer_effort = sum(observer_effort, na.rm = TRUE)) %>%
  tidyr::pivot_wider(names_from = method, values_from = observer_effort)

tot <- t(colSums(dfr[,2:5], na.rm = TRUE))
dfr <- rbind(as.data.frame(dfr), c("Total", tot))
dfr <- rbind(as.data.frame(dfr), c("Rate per year", tot / n_fishing_years))

dfr$BLL   <- sprintf("%.1f", as.numeric(dfr$BLL))
dfr$SLL   <- sprintf("%.1f", as.numeric(dfr$SLL))
dfr$SN    <- sprintf("%.1f", as.numeric(dfr$SN))
dfr$Trawl <- sprintf("%.1f", as.numeric(dfr$Trawl))

pandoc.table(dfr,
             caption = tab_cap,
             row.names = FALSE, 
             emphasize.strong.rows = c(nrow(dfr) - 1, nrow(dfr)),
             missing = "-",
             split.table = Inf, 
             justify = c('left', rep('right', 4)), 
             col.names = c("Fishing year", names(dfr)[2:ncol(dfr)]))
```

```{r, results = "asis", size = "scriptsize"}
tab_cap <- tab_nums(name = "tab:effort_t", caption = paste("Total fishing effort by method and fishing year between ", min(fishing_years)," and ", max(fishing_years), ". Units are of a thousand hooks for bottom longline (BLL) and surface longline (SLL), km of net for set net (SN), and number of tows for trawl."))

dfr <- dat %>% as.data.frame() %>%
  group_by_at(c("fishing_year", "method")) %>% summarise(effort = sum(effort, na.rm = TRUE)) %>%
  tidyr::pivot_wider(names_from = method, values_from = effort)

tot <- t(colSums(dfr[,2:5], na.rm = TRUE))
dfr <- rbind(as.data.frame(dfr), c("Total", tot))
dfr <- rbind(as.data.frame(dfr), c("Rate per year", tot / n_fishing_years))

dfr$BLL   <- sprintf("%.1f", as.numeric(dfr$BLL))
dfr$SLL   <- sprintf("%.1f", as.numeric(dfr$SLL))
dfr$SN    <- sprintf("%.1f", as.numeric(dfr$SN))
dfr$Trawl <- sprintf("%.1f", as.numeric(dfr$Trawl))

pandoc.table(dfr,
             caption = tab_cap,
             row.names = FALSE, 
             emphasize.strong.rows = c(nrow(dfr) - 1, nrow(dfr)),
             missing = "-",
             split.table = Inf, 
             justify = c('left', rep('right', 4)), 
             col.names = c("Fishing year", names(dfr)[2:ncol(dfr)]))
```

```{r}
# Create group to match RasterBrick
dat$group <- format(dat$start_date, "%b")
stopifnot(all(!is.na(dat$group)))
```

```{r}
observer_dat <- dat %>% filter(observer_effort > 0)

# create sp data frame
observer_dat_sp <- sp::SpatialPointsDataFrame(coords = cbind(observer_dat$start_lon, observer_dat$start_lat), data = as.data.frame(observer_dat)[,c("fishing_year", "sra_species", "group", "method", "observer_effort", "n_captures")], proj4string = sp::CRS("+proj=longlat"))
```


```{r, results = 'hide'}
# not sure what's going on here
source("C:/Rpackages/sraInputsBio/R/plot_sra_maps.R")

# calculate observed density and overlap
observed_density <- matrix(nrow = nrow(observer_dat), ncol = length(species))
colnames(observed_density) <- as.character(species)
observed_density[] <- NA

observed_overlap <- observed_density

out <- list()

for (spp in species) {
  
  do.call(data, list(paste0("densities_", spp), package = "sraInputsBio"))
  
  r <- get(paste0("densities_", spp))
  
  stopifnot(rgdal::compare_CRS(raster::crs(r), raster::crs(observer_dat_sp)))
  
  dfr <- get_density(observer_dat_sp, r, effort_name = "observer_effort", group_name = "group")
  
  observed_density[, spp] <- dfr@data$density[,1]
  observed_overlap[, spp] <- dfr@data$density_overlap[,1]
  
  # plot density overlap by model
  rr <- lapply(c("BLL", "SLL", "SN", "Trawl"), function(x) raster::rasterize(subset(dfr, method == x), r, field = "density_overlap"))
  names(rr) <- c("BLL", "SLL", "SN", "Trawl")
  
  rr <- lapply(rr, raster::aggregate, fact = 10)
  
  rr <- raster::stack(rr)

  #gg <- plot_sra_maps(rr)
  #
  #odat_sp <- observer_dat_sp %>% as.data.frame() %>% filter(n_captures > 0) %>% rename(x = coords.x1, y = coords.x2) %>% group_by(x, y, method) %>% summarise(value = sum(n_captures, na.rm = TRUE)) %>% rename(group = method)
  #
  #out[[spp]] <- gg + geom_point(data = odat_sp, aes(x, y, size = value), col = "red", shape = 21, stroke = 0.1) + guides(size = "none")
  #
  out[[spp]] <- knit_child('make_inputs_species.Rmd', envir = new.env())
}

save(observed_overlap, file = file.path("../data", "observed_overlap.rda"))
save(observed_density, file = file.path("../data", "observed_density.rda"))
```

```{r}
rm(observed_density)
```

```{r}
# create sp data frame
total_effort_sp <- sp::SpatialPointsDataFrame(coords = cbind(dat$start_lon, dat$start_lat), data = as.data.frame(dat)[,c("fishing_year", "group", "method", "effort")], proj4string = sp::CRS("+proj=longlat"))
```

```{r, results = 'hide'}
# calculate total density and overlap
total_density <- matrix(nrow = nrow(dat), ncol = length(species))
colnames(total_density) <- as.character(species)
total_density[] <- NA

total_overlap <- total_density

for (spp in species) {
  
  do.call(data, list(paste0("densities_", spp), package = "sraInputsBio"))
  
  r <- get(paste0("densities_", spp))
  
  stopifnot(rgdal::compare_CRS(raster::crs(r), raster::crs(total_effort_sp)))
  
  dfr <- get_density(total_effort_sp, r, effort_name = "effort", group_name = "group")
  
  total_density[, spp] <- dfr@data$density[,1]
  total_overlap[, spp] <- dfr@data$density_overlap[,1]
}

save(total_overlap, file = file.path("../data", "total_overlap.rda"))
save(total_density, file = file.path("../data", "total_density.rda"))
```

```{r}
rm(total_density)
```

```{r}
# join overlap
overlap_1 <- observer_dat %>% as.data.frame() %>%
  select(method, id_method, id_fishing_group, month) %>% 
  cbind(., observed_overlap) %>%
  tidyr::gather("species", "overlap", all_of(species))

stopifnot(nrow(overlap_1) == nrow(observer_dat) * length(species))

# sum overlap
overlap_o <- overlap_1 %>%
  group_by(species, method, id_method, id_fishing_group, month) %>%
  summarise(overlap = sum(overlap, na.rm = TRUE)) %>%
  ungroup() 

# get species groups
overlap_o <- overlap_o %>% 
  left_join(species_groups, by = c('species' = 'code_sra')) %>%
  mutate(month = as.integer(month), id_method = as.integer(id_method), id_fishing_group = as.integer(id_fishing_group), id_cryptic_mortality_group = as.integer(id_cryptic_mortality_group)) %>%
  select(species, method, id_method, month, id_fishing_group, id_species, id_species_group, id_cryptic_mortality_group, overlap)

# get captures
cap_alive <- observer_dat %>% as.data.frame() %>% mutate_captures_status("alive") %>% aggregate_captures(c("id_method", "id_fishing_group", "month"), species) %>% rename(species = group)
cap_dead  <- observer_dat %>% as.data.frame() %>% mutate_captures_status("dead")  %>% aggregate_captures(c("id_method", "id_fishing_group", "month"), species) %>% rename(species = group)
cap       <- full_join(cap_alive, cap_dead, by = c("species", "id_method", "id_fishing_group", "month"), suffix = c("_alive", "_dead"))
cap       <- cap %>% mutate(n_captures = n_captures_alive + n_captures_dead)

cap_sum <- cap %>% group_by(id_method, id_fishing_group, species, month) %>% summarise(n_captures_alive = sum(n_captures_alive), n_captures_dead = sum(n_captures_dead), n_captures = sum(n_captures))

# join overlap and captures
overlap_o <- overlap_o %>%
  left_join(cap_sum, by = c("id_method", "id_fishing_group", "species", "month")) %>%
  tidyr::replace_na(replace = list(n_captures_alive = 0, n_captures_dead = 0))

# remove zero overlap values with no captures
overlap_o <- overlap_o %>% filter(overlap > 0.0 | n_captures > 0)

# checks
stopifnot(isTRUE(all.equal(sum(observed_overlap, na.rm = TRUE), sum(overlap_o$overlap))))
stopifnot(sum(overlap_o$n_captures) == sum(cap$n_captures))

# if captures but no overlap set overlap to small value
overlap_o <- overlap_o %>% mutate(overlap = case_when(overlap == 0.0 & n_captures > 0 ~ 1e-5, TRUE ~ overlap))

# tidy up
rm(observed_overlap)
```

\newpage\clearpage

```{r, results = 'asis', size = 'scriptsize'}
tab_cap <- tab_nums(name = "tab:overlap_o", caption = paste("Observed overlap by species for bottom-longline (BLL), surface-longline (SLL), set net (SN), and trawl fishing methods between ", min(fishing_years), " and ", max(fishing_years), "."))

tab <- overlap_o %>%
  group_by(id_species, species, method) %>%
  summarise(overlap = sum(overlap, na.rm = TRUE)) %>%
  full_join(species_groups, by = c('id_species', 'species' = 'code_sra')) %>%
  select(id_species, species, common_name, method, overlap) %>%
  tidyr::pivot_wider(names_from = method, values_from = overlap, values_fill = list(overlap = 0)) %>%
  ungroup() %>%
  mutate(Total = BLL + SLL + SN + Trawl, Total = ifelse(Total == 0, NA, Total)) %>%
  mutate(BLL = ifelse(BLL == 0, NA, BLL), SLL = ifelse(SLL == 0, NA, SLL), SN = ifelse(SN == 0, NA, SN), Trawl = ifelse(Trawl == 0, NA, Trawl)) %>%
  arrange(as.numeric(id_species)) %>%
  select(-id_species, -species)
if ('NA' %in% colnames(tab)) tab <- tab %>% select(-'NA')
tab <- rbind(tab, data.frame("common_name" = "Total", as.data.frame(t(colSums(tab[,2:6], na.rm = TRUE)))))
tab <- rbind(tab, data.frame("common_name" = "Total per year", as.data.frame(t(colSums(tab[,2:6], na.rm = TRUE) / n_fishing_years))))
tab <- tab %>%
  mutate(BLL = sprintf("%.3f", BLL), SLL = sprintf("%.3f", SLL), SN = sprintf("%.3f", SN), 
         Trawl = sprintf("%.3f", Trawl), Total = sprintf("%.3f", Total)) %>%
  mutate(BLL = ifelse(BLL == "NA", NA, BLL), SLL = ifelse(SLL == "NA", NA, SLL), 
         SN = ifelse(SN == "NA", NA, SN), Trawl = ifelse(Trawl == "NA", NA, Trawl), Total = ifelse(Total == "NA", NA, Total))

pandoc.table(tab,
             caption = tab_cap,
             emphasize.strong.rows = c(nrow(tab) - 1, nrow(tab)),
             emphasize.strong.cols = ncol(tab),
             missing = "-",
             row.names = FALSE, 
             split.tables = Inf,
             justify = c('left', rep('right', 5)), 
             col.names = c("Common name", names(tab)[2:ncol(tab)]))
```



```{r}
# join total effort to overlap
overlap_1 <- dat %>% as.data.frame() %>%
  select(method, id_method, id_fishing_group, month) %>% 
  cbind(., total_overlap) %>%
  tidyr::gather("species", "overlap", all_of(species))

stopifnot(nrow(overlap_1) == nrow(dat) * length(species))

# join to overlap and sum
overlap_t <- overlap_1 %>%
  group_by(species, method, id_method, id_fishing_group, month) %>%
  summarise(overlap = sum(overlap, na.rm = TRUE)) %>%
  ungroup()

# get species groups
overlap_t <- overlap_t %>% 
  left_join(species_groups, by = c('species' = 'code_sra')) %>%
  mutate(month = as.integer(month), id_method = as.integer(id_method), id_fishing_group = as.integer(id_fishing_group), id_cryptic_mortality_group = as.integer(id_cryptic_mortality_group)) %>%
  select(species, method, month, id_method, id_fishing_group, id_species, id_species_group, id_cryptic_mortality_group, overlap)

# check
stopifnot(isTRUE(all.equal(sum(total_overlap, na.rm = TRUE), sum(overlap_t$overlap))))

# if zero commercial overlap but positive observed overlap replace with observed overlap
overlap_t <- full_join(overlap_t, select(overlap_o, species:overlap, n_captures), by = c("species", "method", "id_method", "month", "id_fishing_group", "id_species", "id_species_group", "id_cryptic_mortality_group"))

overlap_t <- overlap_t %>% mutate(overlap = case_when(is.na(overlap.x) ~ overlap.y, 
                                                        overlap.x == 0.0 & overlap.y > 0 ~ overlap.y, 
                                                        TRUE ~ overlap.x)) %>% select(-overlap.x, -overlap.y) 
# remove zero overlap values                                                        
overlap_t <- overlap_t %>% filter(overlap > 0)

# tidy up
rm(total_overlap)
```

\newpage\clearpage

```{r, results = 'asis' , size = 'scriptsize'}
tab_cap <- tab_nums(name = "tab:overlap_t", caption = paste("Total overlap by species for bottom-longline (BLL), surface-longline (SLL), set net (SN), and trawl fishing methods between ", min(dat$fishing_year), " and ", max(dat$fishing_year), "."))

tab <- overlap_t %>%
  group_by(id_species, species, method) %>%
  summarise(overlap = sum(overlap, na.rm = TRUE)) %>%
  full_join(species_groups, by = c('id_species', 'species' = 'code_sra')) %>%
  select(id_species, species, common_name, method, overlap) %>%
  tidyr::pivot_wider(names_from = method, values_from = overlap, values_fill = list(overlap = 0)) %>%
  ungroup() %>%
  mutate(Total = BLL + SLL + SN + Trawl, Total = ifelse(Total == 0, NA, Total)) %>%
  mutate(BLL = ifelse(BLL == 0, NA, BLL), SLL = ifelse(SLL == 0, NA, SLL), SN = ifelse(SN == 0, NA, SN), Trawl = ifelse(Trawl == 0, NA, Trawl)) %>%
  arrange(as.numeric(id_species)) %>%
  select(-id_species, -species)
if ('NA' %in% colnames(tab)) tab <- tab %>% select(-'NA')
tab <- rbind(tab, data.frame("common_name" = "Total", as.data.frame(t(colSums(tab[,2:6], na.rm = TRUE)))))
tab <- rbind(tab, data.frame("common_name" = "Total per year", as.data.frame(t(colSums(tab[,2:6], na.rm = TRUE) / length(unique(dat$fishing_year))))))
tab <- tab %>%
  mutate(BLL = sprintf("%.3f", BLL), SLL = sprintf("%.3f", SLL), SN = sprintf("%.3f", SN), 
         Trawl = sprintf("%.3f", Trawl), Total = sprintf("%.3f", Total)) %>%
  mutate(BLL = ifelse(BLL == "NA", NA, BLL), SLL = ifelse(SLL == "NA", NA, SLL), 
         SN = ifelse(SN == "NA", NA, SN), Trawl = ifelse(Trawl == "NA", NA, Trawl), Total = ifelse(Total == "NA", NA, Total))

pandoc.table(tab,
             caption = tab_cap,
             emphasize.strong.rows = c(nrow(tab) - 1, nrow(tab)),
             emphasize.strong.cols = ncol(tab),
             missing = "-",
             row.names = FALSE, 
             split.tables = Inf,
             justify = c('left', rep('right', 5)), 
             col.names = c("Common name", names(tab)[2:ncol(tab)]))
```

\newpage\clearpage
# Species specific inputs

`r paste(out, collapse = '\n')`


```{r}
# CM parameters
load("cm_pars.rda")
```

```{r}
# Import demographic data
#N_BP   <- read.csv("N_BP.csv")
#S_curr <- read.csv("S_curr.csv")
#S_opt  <- read.csv("S_opt.csv")
#A_curr <- read.csv("A_curr.csv")
#P_B    <- read.csv("P_B.csv")

N_BP   <- inputsBio[["N_BP"]]
S_curr <- inputsBio[["S_curr"]]
S_opt  <- inputsBio[["S_opt"]]
A_curr <- inputsBio[["A_curr"]]
P_B    <- inputsBio[["P_B"]]
p_nest <- inputsBio[["p_nest"]]
p_eez  <- inputsBio[["p_EEZ"]]

N_BP   <- N_BP   %>% filter(code_sra %in% species)
S_curr <- S_curr %>% filter(code_sra %in% species)
S_opt  <- S_opt  %>% filter(code_sra %in% species)
A_curr <- A_curr %>% filter(code_sra %in% species)
P_B    <- P_B    %>% filter(code_sra %in% species)

p_eez  <- p_eez  %>% subset(rownames(.) %in% species)
p_nest <- p_nest %>% subset(rownames(.) %in% species)
```


```{r}
# Check that there are always some birds in zone when there are non-zero captures
tab <- tapply(overlap_o$n_captures, list(overlap_o$species, overlap_o$month), sum, na.rm = TRUE)
tab <- tab[match(species_groups$code_sra, rownames(tab)),]
rownames(tab) <- species_groups$code_sra
colnames(tab) <- colnames(p_eez)

tab <- full_join(reshape2::melt(tab, varnames = c("species", "month")), reshape2::melt(p_eez, varnames = c("species", "month")), by = c("species", "month"))
stopifnot(nrow(tab) == length(species) * 12)
tab <- tab %>% filter(!is.na(value.x))

stopifnot(with(tab, !any(value.x > 0 & value.y == 0)))
```

```{r}
# check that p_nest > 0 for breeding season
#stopifnot(all(p_nest[breeding > 0] > 0))

# check that p_nest == 0 for non-breeding season
#stopifnot(all(p_nest[breeding == 0] == 0))
```

```{r}
# MODEL DATA
sra_dat <- list(n_species = length(species),
            n_month = 12,
            n_method = nrow(methods),
            n_species_group = length(unique(species_groups$species_group)),
            n_cm_group = 5,
            n_fishery_group = nrow(fishing_groups), 
            n_fishery_group_m = as.integer(table(fishing_groups$method)),
            species_group_s = species_groups$id_species_group,
            
            cm_group_s = species_groups$id_cryptic_mortality_group,
            n_cm = 48,
            cm_par1 = cm_pars$par1,
            cm_par2 = cm_pars$par2,
            cm_id = cmidx,
            
            idx_method_fg = fishing_groups$id_method,

            n_breeding_pairs_type = ifelse(N_BP$distribution == 'log-uniform', 0, ifelse(N_BP$distribution == 'log-normal', 1, 2)),
            n_breeding_pairs_p1 = N_BP$par1,
            n_breeding_pairs_p2 = N_BP$par2,
            p_breeding_p1 = P_B$par1,
            p_breeding_p2 = P_B$par2,
            age_breeding_current_p1 = A_curr$par1,
            age_breeding_current_p2 = A_curr$par2,
            adult_survival_current_type = ifelse(S_curr$distribution == 'uniform', 0, 1),
            adult_survival_current_p1 = S_curr$par1,
            adult_survival_current_p2 = S_curr$par2,
            adult_survival_opt_type = ifelse(S_opt$distribution == 'uniform', 0, 1),
            adult_survival_opt_p1 = S_opt$par1,
            adult_survival_opt_p2 = S_opt$par2,
            
            p_eez  = p_eez,
            p_nest = p_nest,
             
            n_i             = nrow(overlap_o),
            month_i         = overlap_o$month,
            method_i        = overlap_o$id_method,
            fishery_group_i = overlap_o$id_fishing_group,
            species_i       = overlap_o$id_species,
            species_group_i = overlap_o$id_species_group,
            cm_group_i      = overlap_o$id_cryptic_mortality_group,
            overlap_i       = overlap_o$overlap,
            live_captures_i = overlap_o$n_captures_alive,
            dead_captures_i = overlap_o$n_captures_dead,
            captures_i      = overlap_o$n_captures_alive + overlap_o$n_captures_dead,
                         
            n_j             = nrow(overlap_t),
            month_j         = overlap_t$month,
            method_j        = overlap_t$id_method,
            fishery_group_j = overlap_t$id_fishing_group,
            species_j       = overlap_t$id_species,
            species_group_j = overlap_t$id_species_group,
            cm_group_j      = overlap_t$id_cryptic_mortality_group,
            overlap_j       = overlap_t$overlap,
            
            n_years = n_fishing_years,
            
            psi = 0.5)
```

```{r}
# check all probability distributions are represented in the model code
stopifnot(isTRUE(all(sra_dat$n_breeding_pairs_type %in% 0:2)))
stopifnot(isTRUE(all(sra_dat$adult_survival_current_type %in% 0:1)))
stopifnot(isTRUE(all(sra_dat$adult_survival_opt_type %in% 0:1)))
```

```{r}
# initial values
sra_ini <- list(log_q_intercept = rep(0, sra_dat$n_method),
                log_q_method_raw = rep(0, sra_dat$n_fishery_group - sra_dat$n_method),
                log_q_species_raw = rep(0.0, sra_dat$n_species_group - 1),
                tau = 1.0,
                eps_gs = matrix(0.0, sra_dat$n_fishery_group, sra_dat$n_species),
                beta_live_capture_fg = rep(2.0, sra_dat$n_fishery_group),
                beta_live_capture_sg = rep(0.0, sra_dat$n_species_group),
                n_breeding_pairs_raw1_s = rep(0.5, sra_dat$n_species - sum(sra_dat$n_breeding_pairs_type)),
                n_breeding_pairs_raw2_s = rep(100, sum(sra_dat$n_breeding_pairs_type)),
                p_breeding_raw_s = rep(2.5, sra_dat$n_species))
```

```{r}
# save
save(methods, file = file.path("../data", "methods.rda"))
save(fishing_groups, file = file.path("../data", "fishing_groups_II.rda"))
save(sra_ini, file = file.path("../data", "sra_ini_update_1.rda"))
save(sra_dat, file = file.path("../data", "sra_dat_update_1.rda"))
save(overlap_o, file = file.path("../data", "overlap_o.rda"))
save(overlap_t, file = file.path("../data", "overlap_t.rda"))
```

\newpage


